---
title: "Week 6: Exploratory Data Analysis (EDA)"
date: "Tuesday, September 26, 2023"
date-format: full
#date: last-modified
#date-format: "dddd MMM D, YYYY, HH:mm:ss"
lecnum: 6
categories:
  - "Class Sessions"
bibliography: "../_DSAN5000.bib"
format:
  revealjs:
    cache: false
    footer: "DSAN 5000-<span class='sec-num'>02</span> Week 6: Exploratory Data Analysis"
  html:
    output-file: index.html
    html-math-method: mathjax
metadata-files: 
  - "../_slide-meta.yml"
---


::: {.content-visible unless-format="revealjs"}

<center>
<a class="h2" href="./slides.html" target="_blank">Open slides in new window &rarr;</a>
</center>

:::

# Schedule {.smaller .small-title .crunch-title .crunch-callout data-name="Schedule"}

::: {.hidden}

```{r}
#| label: r-source-globals
source("../_globals.r")
```

```{python}
#| label: python-globals
from IPython.display import Markdown
def disp(df):
  return Markdown(df.to_markdown())
```

:::

Today's Planned Schedule (Section <span class='sec-num'>02</span>):

{{< include ../_components/sched-w06.qmd >}}

# Week 05 Recap {data-stack-name="Recap"}

* NLP
* Tidyverse
* Merging, Reshaping Data

## NLP Recap {.smaller .small-title .crunch-title .crunch-figures}

{{< include ../_components/doc-term-matrix.qmd >}}

## Your NLP Toolbox {.smaller .crunch-title}

* Processes like **lowercasing** and **stemming** allowed the computer to recognize that `text` and `texts` should be counted together in this context, since they refer to the same semantic concept.
* As we learn NLP, we'll develop a "toolbox" of ideas, algorithms, and tasks allowing us to **quantify, clean, and analyzing text data**, where each tool will help us at some level/stage of this analysis:
  * Gathering texts
  * Preprocessing
  * Learning (e.g., estimating parameters for a model) about the texts
  * Applying what we learned to **downstream tasks** we'd like to solve

## The Items In Our Toolbox {.smaller .crunch-title .crunch-ul .crunch-ul-left}

::: {layout="[20,54,26]" layout-valign="top"}

::: {#gathering-text}

<div style="border: 1px solid black !important; padding: 8px !important;">
<center>
**Corpus Construction**
</center>

&bull; **Corpus**: The collection of documents you're hoping to analyze

&bull; Books, articles, posts, emails, tweets, etc.
</div>

:::
::: {#nlp}

<center>
**Corpus/Document Level NLP**
</center>

&bull; **Vocabulary**: The collection of **unique tokens** across **all documents** in your corpus

&bull; **Segmentation**: Breaking a document into parts (paragraphs and/or sentences)

<center>
&darr;<br>**Sentence/Word Level NLP**
</center>

&bull; **Tokenization**: Break sentence into **tokens**

&bull; **Stopword Removal**: Removing non-**semantic** (syntactic) tokens like "the", "and"

&bull; **Stemming**: Naïvely (but quickly) "chopping off" ends of tokens (e.g., plural &rarr; singular)

&bull; **Lemmatization**: Algorithmically map tokens to linguistic roots (slower than stemming)



:::
::: {#downstream}

<div style="border: 1px solid black !important; padding: 8px !important; margin-bottom: 3px;" class="crunch-p">
<center>
**Vectorization**
</center>

Transform textual representation into numeric representation, like the **DTM**
</div>

<center>
&darr;
</center>

<div class="crunch-p" style="border: 1px solid black !important; padding: 8px !important; margin-top: 3px !important;">
<center>
**Downstream Tasks**
</center>

&bull; **Text classification**

&bull; **Named entity recognition**

&bull; **Sentiment analysis**

</div>

:::

:::

## Tidyverse {.smaller}

* Think of data science tasks as involving **pipelines**:

```{dot}
//| fig-height: 1.5
//| echo: false
digraph G {
  rankdir=LR;
  node[label="Raw Data"] raw;
  subgraph cluster_00 {
    label="Data-Processing Pipeline 1"
    node[label="Transformation A\n(select(), filter())"] tr1;
    node[label="Transformation B\n(mutate(), summarize())"] tr2;
    tr1 -> tr2;
  }
  raw -> tr1;
  node [label="Visualization"] viz;
  tr2 -> viz;
}
```

```{dot}
//| fig-height: 1.5
//| echo: false
digraph G {
  rankdir=LR;
  node[label="Raw Data"] raw;
  subgraph cluster_00 {
    label="Data-Processing Pipeline 2"
    node[label="Transformation C\n(select(), filter())"] tr1;
    node[label="Transformation D\n(mutate(), summarize())"] tr2;
    tr1 -> tr2;
  }
  raw -> tr1;
  node [label="      Result     "] viz;
  tr2 -> viz;
}
```

* Tidyverse lets you **pipe output** from one transformation as the **input** to another:

```r
raw_data |> select() |> mutate() |> visualize()
raw_data |> filter() |> summarize() |> check_result()
```

## *Select*ing Columns {.smaller}

`select()` lets you keep only the **columns** you care about in your current analysis:

::: {layout-ncol=2}

```{r}
#| label: tidyverse-select-load-data
library(tidyverse)
table1
```

```{r}
#| label: tidyverse-select
#| code-fold: show
table1 |> select(country, year, population)
```

:::

## *Filter*ing Rows {.smaller}

`filter()` lets you keep only the **rows** you care about in your current analysis:

::: {layout-ncol=2}

```{r}
#| label: tidyverse-filter-year
#| code-fold: show
table1 |> filter(year == 2000)
```

```{r}
#| label: tidyverse-filter-country
#| code-fold: show
table1 |> filter(country == "Afghanistan")
```

:::

## Merging Data {.smaller .crunch-title}

* The task: Analyze relationship between population and GDP (in 2000)
* The data: One dataset on population in 2000, another on GDP in 2000
* Let's get the data **ready for merging** using R

::: columns
::: {.column width="40%"}

```{r}
#| label: tidyverse-filter-select
#| code-fold: show
df <- table1 |>
  select(country, year, population) |>
  filter(year == 2000)
df |> write_csv("assets/pop_2000.csv")
df
```

:::

::: {.column width="60%"}

```{r}
#| label: load-gdp-data
#| code-fold: show
gdp_df <- read_csv("https://gist.githubusercontent.com/jpowerj/c83e87f61c166dea8ba7e4453f08a404/raw/29b03e6320bc3ffc9f528c2ac497a21f2d801c00/gdp_2000_2010.csv")
gdp_df |> head(5)
```

:::

:::

## Selecting/Filtering in Action {.smaller}

```{r}
#| label: clean-gdp-data
#| code-fold: show
gdp_2000_df <- gdp_df |>
  select(`Country Name`,Year,Value) |>
  filter(Year == "2000") |>
  rename(country=`Country Name`, year=`Year`, gdp=`Value`)
gdp_2000_df |> write_csv("assets/gdp_2000.csv")
gdp_2000_df |> head()
```

## Recommended Language: Python {.smaller .crunch-title .crunch-ul}

Pandas provides an easy-to-use `df.merge(other_df)`!

::: columns
::: {.column width="50%"}

```{python}
#| label: py-load-pop-data
#| echo: false
#| output: false
import pandas as pd
from IPython.display import Markdown
pop_df = pd.read_csv("assets/pop_2000.csv")
```

```{python}
#| label: py-load-gdp-data
#| echo: false
gdp_df = pd.read_csv("assets/gdp_2000.csv")
```

<center>
**Left Join**
</center>

```{python}
#| label: py-merge
#| code-fold: show
merged_df = pop_df.merge(gdp_df,
  on='country', how='left', indicator=True
)
Markdown(merged_df.to_markdown())
```

:::
::: {.column width="50%"}

<center>
**Inner join** (&asymp; Intersection ($\cap$))
</center>

```{python}
#| label: py-merge-right
#| code-fold: show
merged_df = pop_df.merge(gdp_df,
  on='country', how='inner', indicator=True
)
Markdown(merged_df.to_markdown())
```

:::
:::

## Reshaping Data {.smaller}

Sometimes you can't merge because one of the datasets looks like the table on the left, but we want it to look like the table on the right

::: columns
::: {.column width="55%"}

In data-cleaning jargon, this dataset is **long** (more than one row per observation)

```{r}
#| label: r-long-data
table2 |> write_csv("assets/long_data.csv")
table2 |> head()
```

:::
::: {.column width="45%"}

In data-cleaning jargon, this dataset is **wide** (one row per obs; usually **tidy**)

```{r}
#| label: r-wide-data
table1 |> write_csv("assets/wide_data.csv")
table1 |> head()
```

:::
:::

## Reshaping Long-to-Wide in Python: `pd.pivot()` {.smaller}

::: columns
::: {.column width="45%"}

```{python}
#| label: py-load-long-data
#| echo: false
#| output: false
long_df = pd.read_csv("assets/long_data.csv")
```

Create unique ID for **wide** version:

```{python}
#| label: py-create-wide-id
#| code-fold: show
long_df['id'] = long_df['country'] + '_' + long_df['year'].apply(str)
# Reorder the columns, so it shows the id first
long_df = long_df[['id','country','year','type','count']]
disp(long_df.head(6))
```

:::
::: {.column width="55%"}

```{python}
#| label: py-long-to-wide
#| code-fold: show
reshaped_df = pd.pivot(long_df,
  index='id',
  columns='type',
  values='count'
)
disp(reshaped_df)
```

:::
:::

## The Other Direction (Wide-to-Long): `pd.melt()` {.smaller}

::: columns
::: {.column width="40%"}

```{python}
#| label: py-read-wide
#| code-fold: show
wide_df = pd.read_csv("assets/wide_data.csv")
disp(wide_df)
```

:::
::: {.column width="60%"}

```{python}
#| label: py-wide-to-long
#| code-fold: show
long_df = pd.melt(wide_df,
  id_vars=['country','year'],
  value_vars=['cases','population']
)
disp(long_df.head(6))
```

:::
:::

## Wide-to-Long in R: `gather()` {.smaller}

::: columns
::: {.column width="40%"}

```{r}
#| label: r-display-wide-for-reshape
#| code-fold: show
table1
```

:::
::: {.column width="60%"}

```{r}
#| label: r-reshape
#| code-fold: show
long_df <- gather(table1,
  key = "variable",
  value = cases,
  -c(country, year)
)
long_df |> head()
```

:::
:::

# Quiz Time!

* [Quiz 3.1 Link <i class='bi bi-box-arrow-up-right ps-1'></i>](https://georgetown.instructure.com/courses/173310/quizzes/201583){target="_blank"}

# Introduction to EDA {data-stack-name="EDA"}

## Exploratory Data Analysis (EDA)

* In contrast to **confirmatory** data analysis

```{dot}
//| echo: false
//| fig-height: 4
digraph G {
  rankdir=LR;
  nodedir=LR;
  //nodesep=0.5;
  ranksep=0.5;
  //overlap=false;
  //forcelabels=true;

  subgraph cluster_clean {
    margin=28
    label=<
      <table border="0">
        <tr><td><font point-size="24">Clean</font></td></tr>
      </table>
    >

    subgraph cluster_import {
      label="&nbsp;"
      penwidth=0
      margin=0
      node[label=<
        <table border="0">
          <tr><td><font point-size="20">Import</font></td></tr>
        </table>
      >] Import;
      node[label=<
        <table border="0">
          <tr><td>R: <font face="Courier New">read_csv()</font></td></tr>
          <tr><td>Python: <font face="Courier New">pd.read_csv()</font></td></tr>
        </table>
      >,penwidth=0] importLibs;
    }

    subgraph cluster_tidy {
      label="&nbsp;"
      penwidth=0
      margin=0
      node[label=<
        <table border="0">
          <tr><td><font point-size="20">Tidy</font></td></tr>
        </table>
      >] Tidy;
      tidyLibraries[label=<
        <table border="0">
          <tr><td>R: <font face="Courier New">tidyverse</font></td></tr>
          <tr><td>Python: <font face="Courier New">Pandas</font></td></tr>
        </table>
      >,penwidth=0]
    }

    Import -> Tidy
  }
  
	
  subgraph cluster_understand {
    label = <
      <table border="0">
        <tr><td><font point-size="24">Understand</font></td></tr>
      </table>
    >
    margin=10
    subgraph cluster_expl {
      label = "Exploratory";
      margin = 18;
      labeljust="center";
      V[label=<
        <table border="0">
          <tr><td><font point-size="20">Visualize</font></td></tr>
        </table>
      >]
      vizLibraries[label=<
        <table border="0">
          <tr><td>R: <font face="Courier New">ggplot2</font></td></tr>
          <tr><td>Python: <font face="Courier New">seaborn</font></td></tr>
        </table>
      >,penwidth=0]
    }
    subgraph cluster_conf {
      node [label=<
        <table border="0">
          <tr><td><font point-size="20">Model</font></td></tr>
        </table>
      >] Model;
      label = "Confirmatory";
      margin = 18;
      confLibraries[label=<
        <table border="0">
          <tr><td>R: <font face="Courier New">e1071</font></td></tr>
          <tr><td>Python: <font face="Courier New">scikit-learn</font></td></tr>
        </table>
      >,penwidth=0]
    }
  }
  node [label=<
    <table border="0">
      <tr><td><font point-size="20">Communicate</font></td></tr>
    </table>
  >] Communicate;
  
  Tidy -> V;
  V -> Model;
  Model -> V;
  Model -> Communicate;
}
```

## Normalization {.smaller}

::: columns
::: {.column width="42%"}

<center>
**Unnormalized World**
</center>

```{r}
#| label: r-gen-test-scores
#| code-fold: true
num_students <- 30
student_ids <- seq(from = 1, to = num_students)
# Test 1
t1_min_pts <- 0
t1_max_pts <- 268.3
t1_score_vals <- round(runif(num_students, t1_min_pts, t1_max_pts), 2)
t1_mean <- mean(t1_score_vals)
t1_sd <- sd(t1_score_vals)
get_t1_pctile <- function(s) round(100 * ecdf(t1_score_vals)(s), 1)
# Test 2
t2_min_pts <- -1
t2_max_pts <- 1.2
t2_score_vals <- round(runif(num_students, t2_min_pts, t2_max_pts), 2)
t2_mean <- mean(t2_score_vals)
t2_sd <- sd(t2_score_vals)
get_t2_pctile <- function(s) round(100 * ecdf(t2_score_vals)(s), 2)
score_df <- tibble(
  id=student_ids,
  t1_score=t1_score_vals,
  t2_score=t2_score_vals
)
score_df <- score_df |> arrange(desc(t1_score))
```

*"I got a 254.47 on the first test!"* 🤩

*"But only a 0.79 on the second"* 😭

```{r}
#| label: r-test-scores-unnormalized
#| echo: false
score_df |> head()
```

:::
::: {.column width="58%"}

<center>
**Normalized World**
</center>

```{r}
#| label: r-normalize-test-scores
score_df <- score_df |>
  mutate(
    t1_z_score = round((t1_score - t1_mean) / t1_sd, 2),
    t2_z_score = round((t2_score - t2_mean) / t2_sd, 2),
    t1_pctile = get_t1_pctile(t1_score),
    t2_pctile = get_t2_pctile(t2_score)
  ) |>
  relocate(t1_pctile, .after = t1_score) |>
  relocate(t2_pctile, .after = t2_score)
```

*"I scored higher than 96.7% of students on the first test!* 🤩

*"And higher than 90% on the second!"* 😎

```{r}
#| label: r-normalized-scores
#| echo: false
score_df |> head()
```

:::
:::

## Scaling {.smaller .crunch-title .crunch-figures .crunch-images .crunch-code .crunch-p}

The percentile places everyone at **evenly-spaced intervals** from 0 to 100:

```{r}
#| label: pctile-numberline
#| fig-height: 2.5
# https://community.rstudio.com/t/number-line-in-ggplot/162894/4
# Add a binary indicator to track "me" (student #8)
score_df <- score_df |>
  mutate(is_me = as.numeric(id == 8))
library(ggplot2)
t1_line_data <- tibble(
  x = score_df$t1_pctile,
  y = 0,
  me = score_df$is_me
)
ggplot(t1_line_data, aes(x,y,col=factor(me))) +
  geom_point(aes(size=g_pointsize)) +
  scale_x_continuous(breaks=seq(from=0, to=100, by=10)) +
  scale_color_discrete(c(0,1)) +
  dsan_theme("half") +
  theme(
    legend.position="none", 
    #rect = element_blank(),
    #panel.grid = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.y=element_blank(),
    panel.spacing = unit(0, "mm"),
    plot.margin = margin(-35, 0, 0, 0, "pt"),
  ) +
  labs(
    x = "Test 1 Percentile"
  ) +
  coord_fixed(ratio = 100)
```

But what if we want to see their **absolute** performance, on a 0 to 100 scale?

```{r}
#| label: scaled-score-numberline
#| fig-height: 2.5
library(scales)
score_df <- score_df |>
  mutate(
    t1_rescaled = rescale(
      t1_score,
      from = c(t1_min_pts, t1_max_pts),
      to = c(0, 100)
    ),
    t2_rescaled = rescale(
      t2_score,
      from = c(t2_min_pts, t2_max_pts),
      to = c(0, 100)
    )
  )
t1_rescaled_line_data <- tibble(
  x = score_df$t1_rescaled,
  y = 0,
  me = score_df$is_me
)
ggplot(t1_rescaled_line_data, aes(x,y,col=factor(me))) +
  geom_point(aes(size=g_pointsize)) +
  scale_x_continuous(breaks=seq(from=0, to=100, by=10)) +
  dsan_theme("half") +
  theme(
    legend.position="none", 
    #rect = element_blank(),
    #panel.grid = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.y=element_blank(),
    panel.spacing = unit(0, "mm"),
    plot.margin = margin(-40, 0, 0, 0, "pt"),
  ) +
  labs(
    x = "Test 1 Score (Rescaled to 0-100)"
  ) +
  coord_fixed(ratio = 100)
```

## Shifting / Recentering  {.crunch-title .crunch-math}

* Percentiles tell us how the students did **in terms of relative rankings**
* Rescaling lets us reinterpret the **boundary points**
* What about with respect to some **absolute baseline**? For example, how well they did **relative to the mean $\mu$**?

$$
x'_i = x_i - \mu
$$

* But we're still "stuck" in units of the test: is $x'_i = 0.3$ (0.3 points above the mean) "good"? What about $x'_j = -2568$ (2568 points below the mean)? How "bad" is this case?

## Shifting and Scaling: The $z$-Score {.crunch-title .crunch-math .crunch-code .crunch-ul .crunch-images .nostretch}

* Enter the $z$-score!

$$
z_i = \frac{x_i - \mu}{\sigma}
$$

* Unit of original $x_i$ values: ?
* Unit of $z$-score: **standard deviations from the mean**!

```{r}
#| label: z-score-numberline
#| fig-height: 2.5
#| fig-width: 10
t1_z_score_line_data <- tibble(
  x = score_df$t1_z_score,
  y = 0,
  me = score_df$is_me
)
ggplot(t1_z_score_line_data, aes(x,y,col=factor(me))) +
  geom_point(aes(size=g_pointsize)) +
  scale_x_continuous(breaks=c(-2,-1,0,1,2)) +
  dsan_theme("half") +
  theme(
    legend.position="none", 
    #rect = element_blank(),
    #panel.grid = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.y=element_blank(),
    plot.margin = margin(-20,0,0,0,"pt")
  ) +
  expand_limits(x=c(-2,2)) +
  labs(
    x = "Test 1 Z-Score"
  ) +
  coord_fixed(ratio = 3)
```

# Lab 5 {data-stack-name="Lab 5"}


